A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2025-05-26 paper/review-responses.tex

\vskip 1em

We are writing to submit a revised version of our manuscript,
GENETICS-2025-307917 
titled
`` A forest is more than its trees: haplotypes and ancestral recombination graphs''.
We appreciate the thoughtful comments by the reviewers.
We've done our best to address all the comments,
as detailed in our Response to Reviewers.

In addition to a Response to Reviewers (in which page/line numbers refer to the revised manuscript file),
we also provide a pdf with differences to the previously submitted version highlighted.

The main changes we've made are:
(1) additional detail and description to guide the reader through some of the sections;
(2) a fix to a bug in the algorithm
(thanks, Reviewer 3; note this did not affect the code, just the paper);
and (3) more description of how extending haplotypes affects efficiency on inferred ARGs.

@@

\begin{point}{\revref}
 I found the discussion around the metric version to be hard to follow and worth expanding.
For example, I believe that the edges of the ARG are what's getting embedded into
$[0, L) \times N \times N$, and the symmetric differences are taken in this space in the obvious way (i.e., without this mapping its unclear what the relative difference of two ARGs is).
There's also a mention of ``dissimilarity'' in this paragraph. I assume that that's the ``symmetrized version of non-matched span'' that was just defined, but it would be good to make that clear. 
\end{point}

\reply{
    Thanks again to the reviewer for carefully thinking about what we wrote,
    which on further examination has turned out to be wrong in several ways (apologies!).
    Please see the updated version that we believe to be correct
    and thus hopefully less confusing. \revref
}

\begin{point}{}
 The performance on inferred ARGs makes clear that this is a difficult problem, and that this haplotype view might be useful conceptually and for methods development, but that as of now we should probably not take these ancestral haplotypes at face value in inferred ARGs. It would be helpful if the authors could provide some intuition for what's going wrong in this case. Presumably the topologies are not inferred well, and many nodes can be extended, which often results in some true and false positives? Another use case I could see would be using this approach to simplify inferred ARGs to make various downstream computations faster (a la \figref{fig:speed_and_edges}). As such, I would be curious to see how the extension algorithm affects the number of edges in inferred ARGs.
\end{point}

\reply{
    We agree on all counts -- in particular,
    it would be helpful to dig into why haplotypes are being inferred wrongly;
    however, this more detailed investigation is beyond the scope of this paper
    (since it's more an investigation into the inference methods).
    More generally: if an inferred ARG is incorrect for certain nodes, our method of extend haplotypes exacerbates this issue as extending an incorrectly inferred node extends this incorrect node span even further; we've added a note to this effect \revref.
    We've also added \figref{fig:inferred_runtime} that shows numbers of edges and runtime
    for the inferred ARGs used in our tests,
    and additional discussion of this point \llname{inferred_runtime}.
    The results are interesting, but detailed investigation is outside the scope of this paper --
    here we are introducing concepts and tools (and, there's plenty to digest);
    while further investigation could depend strongly on the inference method used.
}

### 2025-05-26 paper/writeup.tex

So, given the ARG shown in \figref[A\&B]{fig:conceptual},
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?
See the discussion for more on what this question assumes,
and the connection to parsimony. \revpoint{1}{7}

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between an ancestor and a descendant.
The ancestor and descendant are represented by the parent node $p_e$ and child node $c_e$ respectively \revpoint{2}{1}
over a segment of genome $[\ell_e, r_e)$.
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

\algorithmref{alg:extend} simplifies the full algorithm implemented in software % in \tskit{}
in several ways for the sake of clarity --
for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
We require repeated passes due to the fringe cases of some paths not being mergeable until 
additional structure is added by a future pass of the algorithm. 
We suspect that most of these cases occur for paths with large time differences between the parent and child
nodes, or when a lot of nodes in a path are coalescent, however it is not something we thoroughly investigated
in the theoretical setting.
We instead give some empirical results on the number of passes for the algorithm below.

@@

Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are otherwise unclear as to the equivalency of nodes across sequences.
(For instance, with a simulated tree sequence and one inferred from its genotypes;
nodes in the former represent actual ancestral haplotypes, and in the latter
represent hypothetical ancestors which may or may not resemble the truth.)
Call the two tree sequences $\T_1$ and $\T_2$, which should have the same genome length
and the same set of sample nodes;
in what follows we think of $\T_2$ as the true ARG and $\T_1$ as an inferred ARG. \llabel{truth}
We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
(b) how much of $\T_2$ is found in $\T_1$; and
(c) how much of $\T_1$ is \emph{not} found in $\T_2$.
(Think of these three quantities as the sizes of two relative intersections
and difference between the tree sequences,
thought of vaguely as sets;
so note in particular that switching the roles of $\T_1$ and $\T_2$ gets 
an entirely different set of quantities.) \llabel{asymmetric}
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).
An example of our method is illustrated in \figref{fig:conceptual_discrepancy}.

@@

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{matched span}, $\match(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets. We first can compute the total span of nodes in $T1$ and $T2$ as $\Vert T1\Vert = 46$ and $\Vert T2\Vert = 47$.
        First, we always require that sample nodes match to their identical counterpart. Then we match intermediate nodes by comparing their
        descendent samples across pairs of local trees over the genome. Here we have distinct local tree pairings on the genome segments:
        $[0,1), [1,2), [2,4), [4,5),$ and $[5,6)$.
        Node $4$ has no match on $[4,5)$, matches with node $9$ on $[5,6)$, and matches with node $4$ in $T2$ everywhere else. 
        Thus the maximal mapping for 4 should be to itself since $m(4,4) > m(4,9)$. 
        Nodes $5, 6,$ and $7$ all match with their identical counterpart.
        Lastly, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
        Given the above matching, the inverse matching will match nodes $0$ through $7$, and node $9$ has no match 
        since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.\newline
        Additionally, let's compute $\operatorname{ARF}(T2,T1)$ and $\operatorname{TPR}(T2,T1)$.
        For matched spans, we again see that sample nodes, root 6, and nodes 4, 5, and 7 will match to their identical counterpart
        and have matched nodes spans 6 (for each sample), 6, 4, 4, and 2, respectively.
        Node 9 matches to node 4 with a matched node span of 1. Thus we compute $\operatorname{ARF}(T2,T1)=\frac{41}{47}$. 
        The inverse matched span will match nodes 5, 6, 7 and sample nodes to their identical counterparts. 
        Node 8 in $T1$ had no match from $T2$, so it has no inverse match. Node 4 had two best matches from $T2$: nodes 4 and 9.
        Since 4 shares more span with its identical counter part the inverse match for 4 in $T1$ is node 4 in $T2$. Therefore we compute 
        $\operatorname{TPR}(T2,T1)=\frac{40}{46}$. 
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

A related (but not identical) way to construct a metric on (undated) ARGs
modifies the definitions above so that $\alpha$ is a bijection.
To do this, write $N_1$ and $N_2$ for the nodes of $\T_1$ and $\T_2$ respectively,
and suppose that $\varphi$ is a bijection from $N_1$ to $N_2$.
Since we want to allow nodes to remain unpaired,
for notational convenience suppose that both $N_1$ and $N_2$ have appended an arbitrary
number of ``unrelated'' nodes (since these are not ancestral to any samples they do not affect the metrics).
Then, define $d_\varphi(\T_1,\T_2)$ to be the total span of nodes $n_1 \in N_1$
on which $n_1$ and $\varphi(n_1)$ do not match (i.e., are not ancestral to the same set of samples),
plus the total span for nodes $n_2 \in N_2$ on which $n_2$ and $\varphi^{-1}(n_2)$ do not match.
The quantity $\min_\varphi d_\varphi(\T_1,\T_2)$,
where the minimum is over bijections that preserve the partial ordering induced by the ARG,
defines a metric on undated ARGs.
(Actually, as written it defines a metric on the equivalence classes of undated ARGs
that are obtained by removing any structure not ancestral to any samples.)
To see this, note that
relative ordering and descendant sample sets uniquely determine the (equivalence class of an) ARG,
and that the triangle inequality is satisfied because
$d_{\varphi_{12}}(\T_1,\T_2) + d_{\varphi_{23}}(\T_2,\T_3) \le d_{\varphi_{12}\circ\varphi_{23}}(\T_1,\T_3)$.

This differs from the definitions above because here
we require a strict one-to-one match between nodes in $\T_1$ and $\T_2$
(i.e., $\varphi$ is bijective, while there may be more than one node in $\T_1$
that $\alpha$ matches to the same node in $\T_2$).
We did not require this property above for two reasons:
first, the arguable non-wrongness of additional unary nodes,
and second, this definition makes computation much more difficult.

@@

As seen in \figref{fig:node-spans},
the vast majority of span added by extending haplotypes is correct.
In this example (which is typical),
99\% of all added span is correct;
95\% of nodes have no incorrectly added span; \revpoint{2}{4}
and those incorrectly added spans are nearly always a small fraction of the original span
(\figref[B]{fig:node-spans} shows that incorrectly added span is often a factor of
10--100 smaller than the added span).
The added information is significant:
the algorithm typically increases spans (i.e., lengths of ancestral haplotypes)
by around 50\%.
For instance, \figref[A]{fig:node-spans} \revpoint{3}{19}
shows that tens of megabases have been added to the spans of many nodes;
while comparing \figref[C]{fig:node-spans} to \figref[D]{fig:node-spans} 
shows that much of the span removed by simplify has been replaced --
in those figures, each point is one node;
span after extending (points in D) are much closer to the original full span
(solid blue points forming a line)
than is span after simpmlifying (points in C).
Some of the incorrectly added spans may be because of violations of the parsimony assumption,
while others are due to situations with ambiguous information
(and hence more than one way to extend a haplotype).
It may be possible to better deal with the latter situation,
at the cost of a more complex algorithm. \llabel{bad_spans}

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
As illustrated by~\cite{wong2024general}, there is a significant diversity
in the structures inferred by current methods,
and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
infers ARGs containing unary nodes as a byproduct of its inference algorithm.
% Note this seems pretty weak and is sure to be dinged. We could at least 
% comment on what we might observe with other inference methods? Would we 
% recommend that people extend_edges() for their SINGER or Relate ARGs?
A comparison across other ARG inference methods is left for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
Since our matched spans method breaks ties with time, 
we date the re-inferred ARG using \tsdate.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 4\% (bottom panels),
but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output
with tsinfer adds relatively little span (less than about $1\%$).
Extending haplotypes with \algorithmref{alg:extend} of the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.) 
We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Fig-6}.

@@

@@

>>>>

So, given the ARG shown in \figref[A\&B]{fig:conceptual},
it should be possible to extend the ancestral haplotype represented by node 3
to obtain the ARG shown in \figref[C\&D]{fig:conceptual},
thus adding additional information to the ARG.
This might be surprising,
as intuition from phylogenetics suggests we can only infer
information about the branching points in the tree, not intermediate (unary) nodes.
The goal of this paper is to answer:
How can we do this, and how accurate is the resulting inference?
See the Discussion for more on what this question assumes,
and the connection to parsimony. \revpoint{1}{7}

@@

We work with the \emph{succinct tree sequence} representation of ARGs (henceforth, ``tree sequence''),
to take advantage of the % extensive, stable, and well-documented set of open-source 
tools available in \tskit{} \citep{tskit}, 
and our terminology and notation follows \citet{ralph2020efficiently}.
For our purposes here,
a tree sequence $\T = (N, E)$ contains a set of \emph{nodes} $N$ 
which represent ancestral segments of genome,
and \emph{edges} $E$ which represent relationships between nodes over different regions of the genome.
Each node $n \in N$ has a \emph{time} $t_n$,
which is the amount of time in the past that the individual who carried that segment of genome lived. \revpoint{3}{13}
Some nodes are \emph{samples}, meaning that they represent genome sequences available as data.
% Typically, samples are exclusively leaves of $\T$, however it is not the case that all samples are leaves.
Each edge $e \in E$ describes inheritance between an ancestor and a descendant
over a segment of genome $[\ell_e, r_e)$.
The ancestor and descendant are represented, respectively, by the parent node $p_e$ and child node $c_e$ of the edge. \revpoint{2}{1}
% (The tree sequence data structure typically includes substantially more information, including genotypes, that we will not use in this paper.)
Suppose that the unique elements of the set of left and right edge endpoints
are $0 = a_0 < a_1 < \cdots < a_{n} = L$, where $L$ is the length of the genome.
Using this information, one can construct the sequence of 
trees $\left(T_1,...,T_{n}\right)$ that describe how the nodes are related to each other along the genome:
each $T_k$ is a tree whose nodes are in $N$
and that describes relationships on the half-open interval $[a_{k-1}, a_k)$.
Nodes represent (portions of) ancestral haplotypes, and so
we will use the terms interchangeably.
Not all nodes appear in each tree,
and we say $n \in T_k$ for a node $n$ if the tree $T_k$ describes at least one parent-child relationship
for node $n$.

@@

\algorithmref{alg:extend} simplifies the full algorithm implemented in software % in \tskit{}
in several ways for the sake of clarity --
for instance, the bookkeeping required to keep track of $T_k$ and $T_{k+1}$ is omitted.
Furthermore, as described the algorithm does one left-to-right pass over the tree sequence;
in practice we do repeated passes in both directions until no changes can be made. \revpoint{2}{2}
We require repeated passes because the additional structure added by a given path of the algorithm
may introduce more mergeable paths.
Most of these cases seem to occur for paths
with large time differences between the parent and child nodes,
or when many nodes in a path are coalescent. 
% however it is not something we thoroughly investigated in the theoretical setting.
The number of required passes is in practice small:
see empirical results below.

@@

Now suppose that instead of comparing two tree sequences with the same set of nodes,
we wish to compare two tree sequences for which we know the sample nodes are the same
but are otherwise unclear as to the equivalency of nodes across sequences.
(For instance, with a simulated tree sequence and one inferred from its genotypes;
nodes in the former represent actual ancestral haplotypes, and in the latter
represent hypothetical ancestors which may or may not resemble the truth.)
Call the two tree sequences $\T_1$ and $\T_2$, which should have the same genome length
and the same set of sample nodes;
in what follows we think of $\T_2$ as the true ARG and $\T_1$ as an inferred ARG. \llabel{truth}
We would like to measure (a) how much of $\T_1$ is found in $\T_2$;
(b) how much of $\T_2$ is found in $\T_1$; and
(c) how much of $\T_1$ is \emph{not} found in $\T_2$.
(Think of these three quantities as the sizes of two relative intersections
and difference between the tree sequences,
thought of vaguely as sets;
so note in particular that switching the roles of $\T_1$ and $\T_2$ gets 
an different set of quantities.) \llabel{asymmetric}
Roughly speaking, we first identify matching nodes
as those whose sets of descendant samples agree for the largest span along the genome,
and then compute for how much of their spans do their descendant samples agree (or not).
An example of our method is illustrated in \figref{fig:conceptual_discrepancy}.

@@

\begin{figure}[!ht]
	\begin{center}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_func_method_t1.pdf}
\includegraphics[height=1.5in, width=3in]{figures/discrepancy_function_method_t2.pdf}
    \caption{
        For two tree sequences $T1$ and $T2$ the \emph{matched span}, $\match(T1,T2)$, matches nodes in $T1$ to nodes in $T2$
        based on identical sample sets. We first can compute the total span of nodes in $T1$ and $T2$ as $\Vert T1\Vert = 46$ and $\Vert T2\Vert = 47$.
        First, we always require that sample nodes match to their identical counterpart. Then we match intermediate nodes by comparing their
        descendent samples across pairs of local trees over the genome. Here we have distinct local tree pairings on the genome segments:
        $[0,1), [1,2), [2,4), [4,5),$ and $[5,6)$.
        Node $4$ has no match on $[4,5)$, matches with node $9$ on $[5,6)$, and matches with node $4$ in $T2$ everywhere else. 
        Thus the maximal mapping for 4 should be to itself since $m(4,4) > m(4,9)$. 
        Nodes $5, 6,$ and $7$ all match with their identical counterpart.
        Lastly, node $8$ has no match in $T2$ as there are no nodes in $T2$ with sample set $\left\{1,2,3\right\}$.
        This makes the matched span $\match(T1,T2)=43$ and $\operatorname{ARF}(T1,T2)=\frac{3}{46}$.
        Given the above matching, the inverse matching will match nodes $0$ through $7$, and node $9$ has no match 
        since its only possible match ($4 \in T1$) was not the best match from $T1 \to T2$. 
        This means the inverse matched span $\invmatch(T1,T2)=43$ and $\operatorname{TPR}(T1,T2)=\frac{43}{47}$.\newline
        Additionally, let's compute $\operatorname{ARF}(T2,T1)$ and $\operatorname{TPR}(T2,T1)$.
        Again, the sample nodes, root 6, and nodes 4, 5, and 7 will match to their identical counterpart
        and have matched nodes spans 6 (for each sample), 6, 4, 4, and 2, respectively.
        Node 9 matches to node 4 with a matched node span of 1.
        Thus, $\operatorname{ARF}(T2,T1)=\frac{41}{47}$. 
        The inverse matched span will match nodes 5, 6, 7 and sample nodes to their identical counterparts. 
        Node 8 in $T1$ had no match from $T2$, so it has no inverse match, while
        node 4 had two best matches from $T2$: nodes 4 and 9.
        Since 4 in $T1$ shares more span with the same-numbered node in $T2$ than with 9,
        the inverse match for 4 in $T1$ is node 4 in $T2$. Therefore,
        $\operatorname{TPR}(T2,T1)=\frac{40}{46}$. 
        \label{fig:conceptual_discrepancy}
    }
	\end{center}
\end{figure}

@@

A related (but not identical) way to construct a metric on (undated) ARGs
modifies the definitions above so that $\alpha$ is a bijection.
To do this, write $N_1$ and $N_2$ for the nodes of $\T_1$ and $\T_2$ respectively,
and suppose that $\varphi$ is a bijection from $N_1$ to $N_2$.
Since we want to allow nodes to remain unpaired,
for notational convenience suppose that both $N_1$ and $N_2$ have appended an arbitrary
number of ``unrelated'' nodes (since these are not ancestral to any samples they do not affect the metrics).
Then, define $d_\varphi(\T_1,\T_2)$ to be the total span of nodes $n_1 \in N_1$
on which $n_1$ and $\varphi(n_1)$ do not match (i.e., are not ancestral to the same set of samples),
plus the total span for nodes $n_2 \in N_2$ on which $n_2$ and $\varphi^{-1}(n_2)$ do not match.
The quantity $\min_\varphi d_\varphi(\T_1,\T_2)$,
where the minimum is over bijections that preserve the partial ordering induced by the ARG,
defines a metric on undated ARGs.
(Actually, as written it defines a metric on the equivalence classes of undated ARGs
that are obtained by removing any structure not ancestral to any samples.)
To see this, note that
relative ordering and descendant sample sets uniquely determine the (equivalence class of an) ARG,
and that the triangle inequality is satisfied because
$d_{\varphi_{12}}(\T_1,\T_2) + d_{\varphi_{23}}(\T_2,\T_3) \ge d_{\varphi_{23}\circ\varphi_{12}}(\T_1,\T_3)$.

This metric differs from the definitions above because here
we require a strict one-to-one match between nodes in $\T_1$ and $\T_2$
(i.e., $\varphi$ is bijective, while there may be more than one node in $\T_1$
that $\alpha$ matches to the same node in $\T_2$).
We did not require this property above for two reasons:
first, the arguable non-wrongness of additional unary nodes,
and second, requiring a bijection makes computation much more difficult.

@@

As seen in \figref{fig:node-spans},
the vast majority of span added by extending haplotypes is correct.
In this example (which is typical),
99\% of all added span is correct;
95\% of nodes have no incorrectly added span; \revpoint{2}{4}
and those incorrectly added spans are nearly always a small fraction of the original span
(\figref[B]{fig:node-spans} shows that incorrectly added span is often a factor of
10--100 smaller than the added span).
The added information is significant:
the algorithm typically increases spans (i.e., lengths of ancestral haplotypes)
by around 50\%.
For instance, \figref[A]{fig:node-spans} \revpoint{3}{19}
shows that tens of megabases have been added to the spans of many nodes;
while comparing \figref[C]{fig:node-spans} to \figref[D]{fig:node-spans} 
shows that much of the span removed by simplification has been replaced --
in those figures, each point is one node;
span after extending (points in D) are much closer to the original full span
(solid blue points forming a line)
than is span after simplifying (points in C).
Some of the incorrectly added spans may be because of violations of the parsimony assumption,
while others are due to situations with ambiguous information
(and hence more than one way to extend a haplotype).
It may be possible to improve the behavior of the algorithm in the latter situation,
at the cost of a more complex algorithm. \llabel{bad_spans}

@@

So far, we have demonstrated that there is potentially ample information
in the coalescent-only trees to extend haplotypes.
Does this work with \emph{inferred} ARGs?
As illustrated by~\cite{wong2024general}, there is a significant diversity
in the structures inferred by current methods,
and here we focus on \tsinfer{}~\cite{kelleher2019inferring} which 
infers ARGs containing unary nodes as a byproduct of its inference algorithm.
% Note this seems pretty weak and is sure to be dinged. We could at least 
% comment on what we might observe with other inference methods? Would we 
% recommend that people extend_edges() for their SINGER or Relate ARGs?
A comparison across other ARG inference methods is left for future work.
We simulated ARGs containing unary-spanning haplotypes (as above), then
re-inferred ARGs from the associated genotypes and performed various operations
on the results.
Since our matched spans method breaks ties with time, 
we date the re-inferred ARG using \tsdate.
First, \figref{fig:dissimilarity} shows that \tsinfer{}
has a substantial portion of already--extended haplotypes:
comparing the ``inferred'' (``I''; dotted green) line to the ``inferred-simplified'' (``IS''; dashed green) line
we see that inclusion of these unary spans increases the amount of correctly inferred material
by around 4\% (bottom panels),
but that roughly 20\% of the unary spans in the inferred ARG are incorrect (top panels).
Furthermore, comparing to the inferred-extend(``IE''; red) line, we see that extending the tree sequence output
with tsinfer adds relatively little span (less than about $1\%$).
Extending haplotypes (with \algorithmref{alg:extend}) of the inferred-and-then-simplified ARG
(``ISE''; solid green line)
produces an ARG with both less correct and incorrect span.
Additionally, the ``IE'' and ``ISE'' ARGs 
contain approximately the same number of edges (difference of $\approx 1\%$).
It is also helpful to note that accuracy
(i.e., proportion of the true ARG that is inferred; bottom panels)
greatly increases with larger sample sizes,
possibly due to resolution of polytomies.
(Recall that due to computational constraints,
``correct'' and ``incorrect'' spans are determined here by \emph{exact} match
of subtending samples.) 
We additionally provide values of ``I'', ``IS'', ``IE'', and ``ISE'' in \tabref{tab:Fig-6},
and numbers of edges and runtime for Tajima's D
in \figref{fig:inferred_runtime}.

@@

Another consideration is storage and computational efficiency: \llabel{inferred_runtime}
extending haplotypes reduces the number of edges,
and thus filesize and (usually) runtime for computation.
\figref{fig:speed_and_edges} shows that the case is clear for simulated ARGs,
substantially reducing both size and runtime.
However, \figref{fig:inferred_runtime}
shows the situation is more complex for inferred ARGs:
extending haplotypes on a \tsinfer-produced ARG actually increases runtime somewhat,
but simplifying-then-extending reduces runtime dramatically
while keeping the number of edges roughly the same.
More work is needed to understand how generalizable this is and what the source of these effects are.

@@

\begin{figure}[!hbt]
    \includegraphics{figures/fig_s5}
	\caption{
        Numbers of edges and runtimes for computing Tajima's D
        for the ARGs displayed in \figref{fig:dissimilarity}.
        Grey lines connect each group of points that derive from the same original simulated ARG
        (i.e., each unique combination of sample size and sequence length).
        The label ``orig'' marks the originally simulated ARG.
    }
	\label{fig:inferred_runtime}
\end{figure}
